Management induced changes of soil organic carbon on global croplands

### 2020-12-11 SOCbudgetPaper/introduction.Rmd
mayor update

Soil Organic Carbon (SOC), the amount of organic carbon stored through the earth’s soil, is the largest terrestrial carbon pool, exceeding the carbon in the atmospheric and vegetation pools multiple times (Batjes). As such, even small changes in drivers of SOC may thus lead to substantial shifts in earth carbon cycle and influence the amount of CO2 in the atmosphere (ref. permafrost melting). The specific amount of carbon stored in the soil is, however, uncertain, with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Batjes, 1996; more). The quality of SOC maps has markedly improved in recent decades, along with better understanding of the factors driving the global magnitude, distribution, and dynamics of SOC pools. Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time.
Human intervention, including land cover change and land management, has however added a further driver to SOC change, which alters terrestrial carbon pools in much shorter time scales and is likely one of the most dominant driver of SOC changes on managed land today (cite). Recent studies identify the anthropogenic SOC debt for the first meter of the soil profile at around 116 GtC (Sanderman et al.), compared to previous estimates between 60-130 GtC  (Lal, 2006). Other studies have focused more closely on spatially disaggregation of SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemical processes within SOC dynamics (Hararuk, 20XX). While providing a rough estimate of the order of magnitude of change, these studies lack a detailed consideration of land management, especially of agricultural activities.
Field-scale models (cite Daycent, RothC, Ecosse, C-Tool) are able to capture these land management impacts by using detailed information on crop yield levels, fertilizer inputs and various other on-farm measures. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global extent remains a complex challenge. Our study combines an spatially explicit estimate of agricultural management data on the global level with a 3-pool SOC model parametrized for global croplands. This allows us to estimate SOC stock change factors, as well as organic carbon flow dynamics within the agricultural system. We thereby consider change in SOC caused by historical land cover change as well as of different agricultural management practices, including residue recycling, manure amendments, irrigation, and tillage. We provide the first global, spatially explicit SOC loss maps that considers detailed agricultural management.
This paper will first introduce in the methods section the basic concept of SOC dynamics as applied in this study. We continue with a detailed description of the global gridded management data used here, including crop production levels, residue recycling rates, manure amendments, and the adoption of irrigation and tillage practices. Lastly, we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines. In our results section we focus on the SOC dynamics of global croplands by (1) analysing the spatially explicit distribution and depletion of SOC from 1975 to 2010, (2) comparing impacts of different management effects on SOC emissions and (3) quantify global agricultural carbon flows and stocks to compare the importance of various management aspects. Finally, we discuss and compare our findings – including their implications for SOC model development – and conclude with an outlook on the ability of SOC management to mitigate climate change and contribute to negative greenhouse gas emissions.

>>>>

Soil Organic Carbon (SOC), the amount of organic carbon stored in the Earth’s soil, is the largest terrestrial organic carbon pool. It exceeds the carbon in the atmospheric and vegetation pools multiple times ([@batjes_total_1996]). Even small changes in processes affecting SOC lead therefore to substantial shifts in the terrestrial carbon cycle and influence the amount of CO~2~ in the atmosphere ([@friedlingstein_global_2019], [@minasny_soil_2017]). The specific amount of carbon stored in soils globally is quantified with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile ([@batjes_total_1996], [@sanderman_soil_2017]).

Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time. Recent studies have focused on the evaluation of total SOC stocks of the world as well as on the spatial disaggregation of soil properties such as SOC content ([@batjes_harmonized_2016], [@hengl_soilgrids250m_2017], [@fao_global_2018]). However, these studies often do not include human interventions, like land cover change and agricultural management, in their analysis. Compared to climatic and geological drivers, they alter terrestrial carbon pools over much shorter time scales and are currently one of most dominant drivers of SOC changes on managed land.

The anthropogenic impact can be measured by the SOC debt (also referred to as SOC component of LUC emissions), which is the amount of organic carbon soils have lost under cultivation compared to a potential natural vegetated state. [@sanderman_soil_2017] identified the anthropogenic SOC debt for the first meter of the soil profile due to land cover change at around 116 GtC (37 GtC for the first 30 cm), compared to previous estimates of 60--130 GtC for the first meter ([@lal_world_2001]).

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs) and Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions ([@friedlingstein_global_2019]). While providing estimates of the magnitude of SOC losses due to land cover change, most models lack a detailed consideration of agricultural management. Earlier DGVM and ESM based assessments only considered changes in land cover, but ignored the removal of biomass at harvest ([@strassmann_simulating_2008], [@betts_climate_2015]). BKMs are designed to estimate LUC realted emissons but ignore changes in SOC due to climate change, CO~2~ fertilization, and agricultural management ([@friedlingstein_global_2019], [@houghton_carbon_2012], [@hansis_relevance_2015]).

Managed agricultural systems were introduced in greater detail to DGVMs and ESMs to improve the assessment of the terrestrial carbon balance (e.g. [@bondeau_modelling_2007], [@lindeskog_implications_2013]). [@pugh_simulated_2015] explicitly consider agricultural management in the form of tillage, irrigation and biomass extraction at harvest, but worked with stylized scenarios rather than with historic management data. They also showed the importance of accounting for the land-use history, as many carbon emissions from agricultural soils are caused by historic LUC and the slow decline of SOC under cropland before it reaches a new equilibrium.

In global-scale carbon cycle assessments, management systems are typically represented as spatially explicit patterns that are static in time (e.g. growing seasons ([portmann_mirca2000global_2010]) multiple cropping systems ([@waha_multiple_2020]), irrigation systems ([@jagermeyr_water_2015])) or as stylized scenarios (e.g. [@pugh_simulated_2015], [@lutz_simulating_2019]).

More data sets on spatially explicit agricultural management time series with global coverage become available (e.g. on tillage systems, see [@porwollik_generating_2018], [@prestele_spatially_2018]) and model approaches are increasingly developed to project the dynamics of management systems into the future (e.g. [@iizumi_modeling_2019], [@minoli_modelling_2019]), but have -- to our knowledge -- not found their way into comprehensive assessments of the terrestrial carbon cycle in DGVMs and BKMs.

<!--Recent studies on soil sequestration for net greenhouse gas removal ([@paustian_quantifying_2019], [@smith_how_2020]) therefore called for 'well-integrated data-model frameworks' including high quality management data to quantify and monitor SOC stock changes.-->

Field-scale models (@del_grosso_simulated_2001], [@coleman_simulating_1997],[@smith_estimating_2010], [@taghizadeh-toosi_c-tool_2014]) are able to better account for historic agricultural management if detailed information on crop yield levels, fertilizer inputs and various other on-farm measures is available for the studied sites. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global domain remains a complex challenge.

The objective of our study is to provide the first global, spatially explicit SOC and SOC debt map that considers spatially explicit and time-variant historical agricultural management. To achieve this objective we create and provide a comprehensive dataset of the global gridded management time series data, including crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We simulate SOC stocks, dynamics and the SOC debt for 1975--2010. Using a scenario analysis, we decompose the contribution of different management activities. Our analysis highlights the importance of considering agricultural management for improving estimates of SOC stocks. We conclude, that historic agricultural management let to increasing SOC stocks in the most recent decades.
